On the Bit Complexity of Finding Points in Connected Components of a Smooth Real Hypersurface

### 2020-02-04 main.tex

%% %
%% \subsubsection{Critical points of polynomial mappings}
%% %
%% Let $\phi : V \rightarrow \C^m$ be a polynomial mapping. A \textit{critical point} of $\phi$ is a regular point of $V$ with $d_x \phi(T_xV) \not = \C^m,$ where $d_x \phi$ is the differential of $\phi$ at $x.$ A \textit{critical} value of $\phi$ is the image of a critical point by $\phi$.
%% %
%% \subsection{The zariski-tangent space}
%% %
%% Denote by $\grad_{\xb}(f)$ the evaluation of the gradient vector of $f$ in $\C[X_1,\hdots,X_n]$ at $\xb$ in $\C$.
%% The \textit{Zariski-tangent space} to $V$ at $\xb \in V$ is the vector space $T_{\xb}V$ defined by the equations $\grad_{\xb}(f) \cdot \textit{\textbf{v}}=0$ for all polynomials $f$ that vanish on $V$. For a polynomial system $\textit{\textbf{f}}=(f_1,\hdots,f_N)$ in $\C[X_1,\hdots,X_n]$, denote by $\jac(\textit{\textbf{f}})$ the Jacobian matrix.
%% The following is a direct consequence of \cite[Corollary 16.20]{ECA}.
%% %
%% \begin{proposition}
%% If $V \subset \C^n$ is a $d$-equidimensional algebraic set with ideal $I(V)=\langle f_1,\hdots,f_N \rangle$, then at any point $\textbf{x}$ of $\reg(V),$ $\jac_\textbf{x}(\textbf{f})$ has full rank $n - \dim(V)$ and the kernel of $\jac_\textbf{x}(\textbf{f})$ is equal to $T_\textbf{x}V.$ 
%% \end{proposition}
%% %
%% \subsection{Polar varieties}
%% %
%% %
%% \subsection{Changes of variables}
%
Consider $A \in \C^{n^2}, f \in \C[X_1,\hdots,X_n], \xb \in \mathbb{C}^n,$ and $V \subset \C^n$. We denote by 
$V^A$ the set $\{x^A~|~\xb \in V\}, f^A$ the polynomial $f(A \xb)$ and $\xb^A$ denotes $A^{-1}\xb$. Notice $f^A(\xb^A) = f(A A^{-1}\xb)=f(\xb).$
%
%
%
%
%
%
%
%
\section{Genericity properties}
%
Consider $f \in \mathbb{Q}[X_1,\hdots,X_n]$ with total degree $d$.  
We say that $f$ satisfies $\textbf{H}$ if 
%
\begin{enumerate}
    \item $f$ is squarefree;
    \item $V(f) \cap \R^n$ is smooth.
\end{enumerate}
%
An equidimensional algebraic set $X$ of dimension $r$ is in \textit{Noether position} for $\pi_r$ when the extension 
\[
\C[X_1,\hdots,X_r] \rightarrow \C[X_1,\hdots,X_n]/I(V)
\] 
is injective and integral.  Then, for any $\xb \in \C^r,$ the fiber 
$V \cap \pi_r^{-1}(\xb)$ has dimension zero. For $i\in\{1,\hdots,n\}$, and for a randomly chosen point $p_{i-1} = (\alpha_1,\hdots,\alpha_{i-1}) \in \Q^{i-1}$, we say that $f$ satisfies $\textbf{H}_i^{'}$ if 
%
\begin{enumerate}
\item $\I(\pi_i,V)$ is radical;
\item either $W(\pi_i,V)$ is empty or $(i-1)$-equidimensional;
\item $W(\pi_i,V)$ is smooth;
\item $\langle X_1-\alpha_1,\hdots,X_{i-1}-\alpha_{i-1}, f,  \frac{\pa f}{\pa X_{i+1}},\hdots,\frac{\pa f}{\pa X_n}\rangle $ is radical; 
\item $V(X_1-\alpha_1,\hdots,X_{i-1}-\alpha_{i-1}, f,  \frac{\pa f}{\pa X_{i+1}},\hdots,\frac{\pa f}{\pa X_n})$ is finite;
\item $V(X_1-\alpha_1,\hdots,X_{i-1}-\alpha_{i-1}, f,  \frac{\pa f}{\pa X_{i+1}},\hdots,\frac{\pa f}{\pa X_n})$ is smooth;
\item either $W(\pi_i,V)$ is empty or in Noether position for $\pi_{i-1}$.
\end{enumerate}
%
\begin{theorem}
Suppose that $f \in \Q[X_1,\hdots,X_n]$ satisfies \textbf{H}. Then there exists a Zariski closed set $\D \subset \mathbb{C}^{n^2}$ of degree at most $D^{\textrm{\td}}$ such that, if $\textbf{A}$ is chosen from  $\C^{n^2}-\Delta$ ( $\Q^{n^2}-\Delta$?) then $f^A$ satisfies $\textbf{H}_i^{'}$ for all $i \in \{1,\hdots,n\}.$
\end{theorem}
%
\begin{corollary} 
Fix $S \subset \mathbb{Q}$ with $|S| \geq \epsilon^{-1} D^{\textrm{\td}}$ and $\epsilon > 0$. Then for $A\in S^{n^2}$ chosen randomly, the probability that $f^A$ satisfies $H_i^{'}$ for all $i \in \{1,\hdots,n\}$ is at least $1-\epsilon.$
\end{corollary}
%
%
%
%
\section{Weak Transversality}

Sard's lemma states that the set of critical values of a smooth
function $\R^n \to \R^m$ has measure zero. One can give ``algebraic''
versions of it, for semi-algebraic mappings $\R^n \to \R^m$ as
in~\cite[Chapter~9]{bochnak1998real} or polynomial mappings $\C^n \to
\C^m$ as in~\cite[Chapter~3]{Mumford76}, for which the sets of critical
values are contained in strict semi-algebraic, resp.\ algebraic sets
in the codomain. Thom's weak transversality lemma, as given for
instance in~\cite{demazure2000bifurcations}, generalizes Sard's
lemma. In this section, we consider a particular case of this result
(transversality to a point), and establish a quantitative version of
it. Our presentation is inspired by that in~\cite{TWT}.

@@

Let $p_{n-1} = (x_1,\hdots,x_{n-1}) \in \Q^{n-1}$ be a randomly chosen point. For $i \in \{1,\hdots,n\}$, we want to show that the Jacobian of the system of polynomials

@@

X_1(\xb)-x_1,\hdots,X_{i-1}(\xb)-x_{i-1}, f(\xb),  \frac{\pa f(\xb)}{\pa X_{i+1}},\hdots,\frac{\pa f(\xb)}{\pa X_n}

>>>>

%% %
%% \subsubsection{Critical points of polynomial mappings}
%% %
%% Let $\phi : V \rightarrow \C^m$ be a polynomial mapping. A \textit{critical point} of $\phi$ is a regular point of $V$ with $d_x \phi(T_xV) \not = \C^m,$ where $d_x \phi$ is the differential of $\phi$ at $x.$ A \textit{critical} value of $\phi$ is the image of a critical point by $\phi$.
%% %
%% \subsection{The zariski-tangent space}
%% %
%% Denote by $\grad_{\xb}(f)$ the evaluation of the gradient vector of $f$ in $\C[X_1,\hdots,X_n]$ at $\xb$ in $\C$.
%% The \textit{Zariski-tangent space} to $V$ at $\xb \in V$ is the vector space $T_{\xb}V$ defined by the equations $\grad_{\xb}(f) \cdot \textit{\textbf{v}}=0$ for all polynomials $f$ that vanish on $V$. For a polynomial system $\textit{\textbf{f}}=(f_1,\hdots,f_N)$ in $\C[X_1,\hdots,X_n]$, denote by $\jac(\textit{\textbf{f}})$ the Jacobian matrix.
%% The following is a direct consequence of \cite[Corollary 16.20]{ECA}.
%% %
%% \begin{proposition}
%% If $V \subset \C^n$ is a $d$-equidimensional algebraic set with ideal $I(V)=\langle f_1,\hdots,f_N \rangle$, then at any point $\textbf{x}$ of $\reg(V),$ $\jac_\textbf{x}(\textbf{f})$ has full rank $n - \dim(V)$ and the kernel of $\jac_\textbf{x}(\textbf{f})$ is equal to $T_\textbf{x}V.$ 
%% \end{proposition}

\section{Genericity properties}

Consider $f \in \ZZ[X_1,\hdots,X_n]$ with total degree $d$, and assume
that $f$ is squarefree and that $V(f) \subset \C^n$ is smooth. The key
to the proof of Theorem~\ref{theo:main} is the following quantitative
version of facts we stated above, namely that in generic coordinates,
polar varieties are smooth, equidimensional, and in Noether position.

We recall that an equidimensional algebraic set $X \subset \C^n$ of
dimension $d$ is in \textit{Noether position} for the projection
$\pi_d$ when the extension $\C[X_1,\hdots,X_d] \rightarrow
\C[X_1,\hdots,X_n]/I(X)$ is injective and integral; here, $I(X)
\subset \C[X_1,\dots,X_n]$ is the defining ideal of $X$. In this case,
for any $\xb \in \C^d,$ the fiber $X \cap \pi_d^{-1}(\xb)$ has
dimension zero (so it is finite and not empty).

For $i$ in $\{1,\dots,n-1\}$ and $f$ as above, we will let $\I(i, f)$
denote the sequence of $n-(i-1)$ polynomials $(f, {\pa f}/{\pa
  X_{i+1}},\dots,{\pa f}/{\pa X_n})$. As we pointed out in the
introduction, their zero-set is the $i$th polar variety $W(\pi,V(f))$.
Then, we say that $f$ satisfies $\textbf{H}_i$ if
\begin{enumerate}
\item For any $\xb$ in $W(\pi,V(f))$, the Jacobian matrix
  $\jac(\I(i, f))$ has full rank $n-(i-1)$ at $\xb$.

  {\em By the Jacobian Criterion~\cite[Corollary 16.20]{ECA}, this
  implies that $W(\pi_i,V(f))$ is either empty or $(i-1)$-equidimensional.}

\smallskip

\item $W(\pi_i,V(f))$ is either empty or in Noether position for
  $\pi_{i-1}$.
\end{enumerate}
When this holds, given $\bm \alpha = (\alpha_1,\hdots,\alpha_{i-1})$
in $\C^{i-1}$, we further say that $f$ and $\bm \alpha$ satisfy
$\textbf{H}'_i$ if
\begin{enumerate}
\item the $n$ equations
  $(X_1-\alpha_1,\dots,X_{i-1}-\alpha_{i-1}),\I(i, f))$ generate a
  radical ideal.

  {\em By point (2) above, we know that there are finitely many solutions 
    to these equations.}
\end{enumerate}

Even if $f$ does not satisfy $\textbf{H}_i$, we mentioned that it does
after applying a change of variables. The precise statement is the
following; to state it, we use the following notation. For a matrix
$\mA$ in $\C^{n^2}$ and $g$ in $\C[X_1,\hdots,X_n]$ we write
$g^\mA:=g(\mA \Xb) \in \C[X_1,\dots,X_n]$.

%% , $\xb \in
%% \mathbb{C}^n$ and $X \subset \C^n$. We write $\xb^\mA:=\mA^{-1}\xb$,
%% $X^\mA:=\{x^\mA~|~\xb \in V\}$ and $g^\mA:=g(\mA \Xb) \in
%% \C[X_1,\dots,X_n]$. In particular, we have $g^\mA(\xb^\mA) =g(\xb)$,
%% and $V(f)^\mA=V(f^\mA)$.

\begin{theorem}
  There exists a hypersurface $\D \subset \mathbb{C}^{n^2}$ of degree
  at most $\td$ such that, if $\mA$ is chosen from $\C^{n^2}-\Delta$
  then $\mA$ is invertible and $f^\mA$ satisfies $\textbf{H}_i$ for
  all $i \in \{1,\hdots,n-1\}.$
\end{theorem}
%% %
%% \begin{corollary} 
%%   Fix $S \subset \mathbb{Q}$ with $|S| \geq \epsilon^{-1}
%%   D^{\textrm{\td}}$ and $\epsilon > 0$. Then for $A\in S^{n^2}$ chosen
%%   randomly, the probability that $f^A$ satisfies $H_i^{'}$ for all $i
%%   \in \{1,\hdots,n\}$ is at least $1-\epsilon.$
%% \end{corollary}

We will also work with change of variables with ``generic entries''.
For this, we will introduce $n^2$ new indeterminates $(\frak
A_{i,j})_{1\le i,j \le n^2}$, we will let $\K$ be the field $\C(\frak
A_{i,j})_{1\le i,j \le n^2})$, and use the corresponding matrix $ \fA$
over $\K$. For $g$ as above, we will then define the polynomial 
$g^\fA:=g(\fA \Xb) \in \K[X_1,\dots,X_n]$.

\section{Weak Transversality}

Sard's lemma states that the set of critical values of a smooth
function $\R^n \to \R^m$ has measure zero. One can give ``algebraic''
versions of it, for semi-algebraic mappings $\R^n \to \R^m$ as
in~\cite[Chapter~9]{bochnak1998real} or polynomial mappings $\C^n \to
\C^m$ as in~\cite[Chapter~3]{Mumford76}, for which the sets of
critical values are contained in strict semi-algebraic,
resp.\ algebraic sets in the codomain. Thom's weak transversality
lemma, as given for instance in~\cite{demazure2000bifurcations},
generalizes Sard's lemma. In this section, we consider a particular
case of this result (transversality to a point), and establish a
quantitative version of it; this will allow us to establish the first
item in property ${\bf H}_i$, as well as property ${\bf H}'_i$. Our
presentation is inspired by that in~\cite{TWT}.

@@

Let $p_{n-1} = (\alpha_1,\hdots,\alpha_{n-1}) \in \Q^{n-1}$ be a randomly chosen point. For $i \in \{1,\hdots,n\}$, we want to show that the Jacobian of the system of polynomials

@@

X_1(\xb)-\alpha_1,\hdots,X_{i-1}(\xb)-\alpha_{i-1}, f(\xb),  \frac{\pa f(\xb)}{\pa X_{i+1}},\hdots,\frac{\pa f(\xb)}{\pa X_n}
